Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-08-22 analysis/paper/paper.Rmd
ongoing

The most widely available source of boreal caribou demographic information is from cows fitted with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@decesare_estimating_2012], though genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g. @hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou). Recruitment is estimated from aerial surveys of the composition of groups of animals that include collared animals [@nbckc_nbckc_2022]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020]. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa , and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020]. Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. In Bayesian population models, all available demographic or population data (e.g. abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021]. However, there is very little monitoring data available for some caribou populations [@johnson_science_2020] and there is a risk that lack of local data will be used as a reason to delay conservation action [@boan_climate_2018]. Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships, and we are concerned that models informed only by local demographic data could be used to inflate uncertainty about the status of poorly monitored populations.

@@

Simulated local demographic and covariate data are provided as inputs to the Bayesian population model, including the observed state (dead or alive) of each collared animal in each month ($\hat{I}_{i,t,m}$), the observed number of adult females and calves in the recruitment surveys ($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per collared female in the composition survey $w$, and the measures of anthropogenic disturbance and fire ($A_t, F_t$). Following Eacker et al. [-@eacker_web-based_2019], recruitment rate $R_t$ is estimated from the observed number of calves $\hat{J}_t$ and adult female caribou $\hat{W}_t$ assuming a binomial distribution:
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

@@

Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the _survival_ R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a modelled as a Gaussian
distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
-->

>>>>

The most widely available source of boreal caribou demographic information is from cows fitted with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@decesare_estimating_2012], though genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g. @hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013, @eacker_web-based_2019, and @lamb_effectiveness_2024 for examples of application to caribou). Recruitment is estimated from aerial surveys of the composition of groups of animals that include collared animals [@nbckc_nbckc_2022]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020]. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa , and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020]. Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019;@lamb_effective_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. In Bayesian population models, all available demographic or population data (e.g. abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021]. However, there is very little monitoring data available for some caribou populations [@johnson_science_2020] and there is a risk that lack of local data will be used as a reason to delay conservation action [@boan_climate_2018]. Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019;@lamb_effectiveness_2024] have not thus far included demographic-disturbance relationships, and we are concerned that models informed only by local demographic data could be used to inflate uncertainty about the status of poorly monitored populations.

@@

Local demographic and covariate data provided as inputs to the Bayesian population model include the observed state (dead or alive) of each collared animal in each month ($\hat{I}_{i,t,m}$), the observed number of adult females and calves in the recruitment surveys ($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per collared female in the composition survey $w$, and the measures of anthropogenic disturbance and fire ($A_t, F_t$). Recruitment rate $R_t$ is estimated from the observed number of calves $\hat{J}_t$ and adult female caribou $\hat{W}_t$ assuming a Poisson distribution:
\begin{equation}
\hat{J}_t \sim \text{Poisson}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

@@

Following others [@eacker_web-based_2019;@lamb_effectiveness_2024], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the _survival_ R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a modelled as a Gaussian
distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019,@lamb_effectiveness_2024]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives biased results if there is left censoring, or months without animals at risk. To address this we drop left censored observations and only estimate survival when at least some animals remain at risk by month 12.
\*
see alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. See also Lamb et al, who uses two step KM estimation process for survival like Eacker. Stick with published examples. \*
-->
